Modeling trend in temperature volatility using generalized LASSO

### 2018-05-16 nips-2018/paper.tex
update paper

\attn{Some equations do not have numbering and some have.}

@@

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). First, we note that
this is a generalized LASSO problem
\citep{tibshirani_solution_2011}. The dual of a generalized LASSO with
the objective $f(x)+\lambda \norm{ Dx }_1$ is:

\begin{align}
\min_\nu&\quad f^*(-D^\top\nu) & \mbox{s.t.}&\quad \norm{ \nu }_\infty \le \lambda
\end{align}
 where $f^*(\cdot)$ is the Fenchel conjugate of $f$: $f^*(u)=\max_x
 u^\top x-f(x)$. It is simple to show that 
\begin{equation}
f^*(u)=\sum_t (u_t-1)\log\frac{y_t^2}{1-u_t} + u_t-1.
\label{eq:conj}
\end{equation}
Writing
\begin{equation}
  r_w(v,\mu_1,\mu_2):=
  \begin{bmatrix}
    \nabla f^*(-D^\top v) + D(v-\lambda \one)^\top \mu_1 -
    D(v+\lambda \one)^\top \mu_2\\
    -\mu_1(v-\lambda\one)+\mu_2(v + \lambda\one) -w^{-1}\one
  \end{bmatrix}
  =
  \begin{bmatrix}
  0\\0\end{bmatrix}
\label{eq:resid}
\end{equation}

for $w>0$, where $\mu_1$ and $\mu_2$ are dual variables for the
$\ell_\infty$ constraint, as $w\rightarrow\infty$,
the solution to this nonlinear system reduces to the Karush–Kuhn–Tucker (KKT)
conditions. Therefore, the PDIP method takes Newton steps to solve the
system for a series of increasing values of $w$. \autoref{alg:pdip}
provides the details.

\begin{algorithm}[tb]
  \caption{PDIP for $\ell_1$ variance estimation}
  \label{alg:pdip}
  \begin{algorithmic}
    \REQUIRE $\lambda>0$, $w>0$, $\nu\leftarrow 0$, $\mu_1\leftarrow
    0$, $\mu_2\leftarrow 0$, $J\in\mathbb{Z}^+$, $\{w_k\}$
    \COMMENT{Initialization} 
    \FOR[Central path]{$k=1,2,\ldots$}
    \FOR[Newton updates]{$j=1,\ldots,J$}
    \STATE Solve $A [\Delta\nu\; \Delta\mu_1\; \Delta\mu_2]^\top=r_{w_k}$ to find the search direction.
    \STATE $\quad\quad A$ is the Jacobian of $r_w$ in \eqref{eq:resid}.
    \STATE Update $[\nu^{j+1}\; \mu^{j+1}_1\; \mu^{j+1}_2] \leftarrow
    [\nu^{j}\; \mu^{j}_1\; \mu^{j}_2] + [\Delta\nu\; \Delta\mu_1\;
    \Delta\mu_2]$. 
    \ENDFOR
   \ENDFOR
   \RETURN $h=\log\frac{y^2}{1+D^\top\nu}$ \attn{Is this right?
     Explicit form of $A$}
  \end{algorithmic}
\end{algorithm}

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend
filtering on graphs~\citep{wang_trend_2016} which penalizes the difference
between the estimated values of the signal on the connected
nodes. As before, we can write
\eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times
S$ vector and $D$ is replaced by $D_{TS} \in \mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot (T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and spatial constraints, respectively\footnote{$N_s$ is obtained by counting the number of unique constraints at each location and at all times.}. Then, as we
have two different tuning parameters for the temporal and spatial
components, we write $\Lambda =\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading to \footnote{Throughout the paper, we use $|x|$ for both scalars and vectors. For vectors we use this to denote a vector obtained by taking the absolute value of each entry of $x$.}:
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -y\given h)+ \Lambda^\top | D_{TS}h |
\label{eq:l1tf_var_st_mat}
\end{equation}
% where $\Lambda = \mathbf{I}_{Tn_rn_c}\left[\lambda_1\one_T^\top,\;
%   \lambda_2\one_{n_rn_c}^\top\right]$ and 
% $D_{total}^t=[D_{t}^t|D_{s}^t]$, where $n_t$ and $n_s$ are the number
% of rows of $D_{t}$ and $D_{s}$, respectively (see below). The matrix
% $D_t$ is the following block-diagonal matrix and corresponds to the
% temporal penalty: 
% \begin{equation}
% D_t=\begin{bmatrix}
% D &  & \\ 
%  & \ddots & \\
%  &  & D
% \end{bmatrix}
% \label{eq:d_t_matrix}
% \end{equation}
%  where $D$ was defined in \eqref{eq:d_matrix}. The number of the
%  diagonal blocks is equal to the grid size $n_r \times n_c$. Each row
%  of the matrix $D_s$ corresponds to one spatial constraint in
%  \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to
%  $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to
%  $|h_{11t}-h_{12t}|$, and so on. 
% \attn{Detailed discussion of the PDIP to here}

@@

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $ST$ columns. For a $1^\circ\times
1^\circ$ grid over the
entire northern hemisphere and daily data over 10 years, we have
$n_r=90$, $n_c=180$, $T=3650$ and so $D_{ST}$ has approximately
$10^8$ columns and $10^8$
rows. In each
step of \autoref{alg:pdip}, we need to solve a linear system of
equations in $A$ which depends on $D_{ST}^\top D_{ST}$ (see \citep{boyd_convex_2004}
equation 11.54). Therefore, 
applying the PDIP directly is infeasible for our data.\footnote{We note that this
  is a highly structured and sparse matrix, but, unlike trend
  filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

In the next section, we develop two ADMM algorithms for solving this
problem efficiently. The first casts the problem as a
so-called consensus optimization problem~\citep{boyd_distributed_2011}
which solves smaller sub-problems using PDIP and then recombines the results. The
second uses proximal methods to avoid matrix inversions. 
\attn{We note that stochastic gradient descent could be used but...}

@@

\attn{I'm not sure if we want to devote some space to write this as a separate algorithm. The steps are those mentioned in \ref{eq:ADMM_steps} and the x-update is now explained in the appendix. So we may want to remove algorithm \ref{alg:ADMM}.}

@@

In this appendix we provide more details on how to solve the optimization problem \ref{eq:l1tf_var}

>>>>

\attn{Arash: Some equations do not have numbering and some have.}

\attn{Arash: fix appendicies and the references to them.}

@@

As with~\eqref{eq:l1tf}, one can solve~\eqref{eq:l1tf_var} using the
PDIP algorithm (as in, e.g.,
\texttt{cvxopt}~\citep{andersen_cvxopt:_2013}). In each iteration of PDIP we need to compute a \textit{search direction} by taking a Newton step on a system of nonlinear equations. Appendix B provides details on how this computation can be performed for the objective function \ref{eq:l1tf_var}.

\attn{Arash: I think we should remove Algorithm 1. There are other steps in PDIP which involve adapting the algorithm parameter $w$ and doing Backtracking line search. Also there is only one loop in PDIP.}

\begin{comment}
\begin{algorithm}[tb]
  \caption{PDIP for $\ell_1$ variance estimation}
  \label{alg:pdip}
  \begin{algorithmic}
    \REQUIRE $\lambda>0$, $w>0$, $\nu\leftarrow 0$, $\mu_1\leftarrow
    0$, $\mu_2\leftarrow 0$, $J\in\mathbb{Z}^+$, $\{w_k\}$
    \COMMENT{Initialization} 
    \FOR[Central path]{$k=1,2,\ldots$}
    \FOR[Newton updates]{$j=1,\ldots,J$}
    \STATE Solve $A [\Delta\nu\; \Delta\mu_1\; \Delta\mu_2]^\top=r_{w_k}$ to find the search direction.
    \STATE $\quad\quad A$ is given in \ref{eq:delta_r}
    \STATE Update $[\nu^{j+1}\; \mu^{j+1}_1\; \mu^{j+1}_2] \leftarrow
    [\nu^{j}\; \mu^{j}_1\; \mu^{j}_2] + [\Delta\nu\; \Delta\mu_1\;
    \Delta\mu_2]$. 
    \ENDFOR
   \ENDFOR
   \RETURN $h=\log\frac{y^2}{1+D^\top\nu}$ 
   \\
   \attn{Is this right? Explicit form of $A$}\\
   \attn{Arash: see appendix A}
  \end{algorithmic}
\end{algorithm}
\end{comment}

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend
filtering on graphs~\citep{wang_trend_2016} which penalizes the difference
between the estimated values of the signal on the connected
nodes. As before, we can write
\eqref{eq:l1tf_var_st} in matrix form where $h$ is an $T\times
S$ vector and $D$ is replaced by $D_{TS} \in \mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot (T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and spatial constraints, respectively\footnote{$N_s$ is obtained by counting the number of unique constraints at each location and at all times.}. Then, as we
have two different tuning parameters for the temporal and spatial
components, we write $\Lambda =\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading to \footnote{Throughout the paper, we use $|x|$ for both scalars and vectors. For vectors we use this to denote a vector obtained by taking the absolute value of each entry of $x$.}:
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{TS}h |
\label{eq:l1tf_var_st_mat}
\end{equation}
% where $\Lambda = \mathbf{I}_{Tn_rn_c}\left[\lambda_1\one_T^\top,\;
%   \lambda_2\one_{n_rn_c}^\top\right]$ and 
% $D_{total}^t=[D_{t}^t|D_{s}^t]$, where $n_t$ and $n_s$ are the number
% of rows of $D_{t}$ and $D_{s}$, respectively (see below). The matrix
% $D_t$ is the following block-diagonal matrix and corresponds to the
% temporal penalty: 
% \begin{equation}
% D_t=\begin{bmatrix}
% D &  & \\ 
%  & \ddots & \\
%  &  & D
% \end{bmatrix}
% \label{eq:d_t_matrix}
% \end{equation}
%  where $D$ was defined in \eqref{eq:d_matrix}. The number of the
%  diagonal blocks is equal to the grid size $n_r \times n_c$. Each row
%  of the matrix $D_s$ corresponds to one spatial constraint in
%  \eqref{eq:l1tf_var_st}. For example, the first $T$ rows correspond to
%  $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to
%  $|h_{11t}-h_{12t}|$, and so on. 
% \attn{Detailed discussion of the PDIP to here}

@@

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have $3Tn_rn_c-2n_rn_c-Tn_r$ rows and $ST$ columns. For a $1^\circ\times 1^\circ$ grid over the entire northern hemisphere and daily data over 10 years, we have $n_r=90$, $n_c=180$, $T=3650$ and so $D_{ST}$ has approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP algorithm, we need to solve a linear system of equations in $A$ which depends on $D_{ST}^\top D_{ST}$ (see appendix A and B). Therefore, applying the PDIP directly is infeasible for our data.\footnote{We note that this is a highly structured and sparse matrix, but, unlike trend filtering alone, it is not banded. We are unaware of general linear algebra techniques for inverting such matrix, despite our best efforts.}

In the next section, we develop two ADMM algorithms for solving this
problem efficiently. The first casts the problem as a
so-called consensus optimization problem~\citep{boyd_distributed_2011}
which solves smaller sub-problems using PDIP and then recombines the results. The
second uses proximal methods to avoid matrix inversions.

\attn{We note that stochastic gradient descent could be used but...}

\attn{Arash: The primal objective is not differentiable and so we cannot use SGD. The dual has constriants and so the plane SGD cannot be used. We can use projected SGD but I think it would converge slowly. Since other papers on LASSO have not considered SGD I don't think we need to justify why we did not use it.}

@@

\attn{I'm not sure if we want to devote some space to write this as a separate algorithm. The steps are those mentioned in \ref{eq:ADMM_steps} and the x-update is now explained in the appendix. So we may want to remove algorithm \ref{alg:ADMM}. }

@@

In this appendix we provide more details on how to solve the optimization problem \ref{eq:l1tf_var} using PDIP. The objective function is convex but not differentiable. Therefore, to be able to use PDIP we first need to derive the dual of this problem. We note that this is a generalized LASSO problem \citep{tibshirani_solution_2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \norm{ Dx }_1$ is:

\begin{align}
\min_\nu&\quad f^*(-D^\top\nu) & \mbox{s.t.}&\quad \norm{ \nu }_\infty \le \lambda
\end{align}

where $f^*(\cdot)$ is the Fenchel conjugate of $f$: $f^*(u)=\max_x u^\top x-f(x)$. It is simple to show that

\begin{equation}
f^*(u)=\sum_t (u_t-1)\log\frac{y_t^2}{1-u_t} + u_t-1.
\label{eq:conj}
\end{equation}

Each iteration of PDIP involves computing a search direction by taking a Newton step for the system of nonlinear equations $r_w(v,\mu_1,\mu_2)=0$, where $w>0$ is a parameter and

\begin{equation}
  r_w(v,\mu_1,\mu_2):=
	\begin{bmatrix}
	r_{dual}\\
	r_{cent}	
	\end{bmatrix}=
  \begin{bmatrix}
    \nabla f^*(-D^\top v) + \mu_1 - \mu_2\\
    -\mu_1(v-\lambda\one)+\mu_2(v + \lambda\one) -w^{-1}\one
  \end{bmatrix}
\label{eq:resid}
\end{equation}

for $w>0$, where $\mu_1$ and $\mu_2$ are dual variables for the $\ell_\infty$ constraint. Let $A=[\nabla r_{dual}^\top , \nabla r_{cent}^\top]^\top$. The newton step takes the following form

\begin{equation}
r_w(v,\mu_1,\mu_2)+A
\begin{bmatrix}
	\nabla v\\
	\nabla \mu_1\\
	\nabla \mu_2	
	\end{bmatrix}= 0
\label{eq:newton_step}
\end{equation}

We have:

\begin{equation}
A=
\begin{bmatrix}
	\nabla^2 f^*(-D^\top v) & I & -I\\
	-\mathbf{diag(\mu_1)}\one & -v+\lambda\one & \mathbf{0}\\
	\mathbf{diag(\mu_2)}\one & v+\lambda\one & \mathbf{0}
	\end{bmatrix}
\label{eq:delta_r}
\end{equation}

Therefore, to perform the Newton step we need to compute $\nabla f^*(-D^\top v)$ and $\nabla^2 f^*(-D^\top v)$. It is straightforward to show that

\begin{align}
\nabla f^*(-D^\top v) = -\nabla_u f^*(u) D^\top ,\quad u=-D^\top v, \quad (\nabla_u f^*(u))_j=\log\bigg(\frac{y_j^2}{1-u_j}\bigg) \\
\nabla^2 f^*(-D^\top v)=D\nabla_u^2 f^*(u)D^\top, \quad (\nabla_u^2 f^*(u))_j=\mathbf{diag}\bigg(\frac{1}{1-u_j}\bigg)
\end{align}
